A* with Fast and Tight Heuristic in Road Networks

### 2020-08-13 paper/ch_potentials.tex
improve pic
some cleanup in story and context

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
In contrast, for road networks there also exist hierarchical speedup techniques, which do not need a heuristic.
They achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use such as Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
We call our technique CH-Potential.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

@@

\centering
\includegraphics[width=\columnwidth]{fig/searchspace.png}

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
This is the tightest potential achievable during preprocessing.
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, it is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

In this section, we first describe Contraction Hierarchies, then PHAST, and finally CH-Potentials.

@@

A CH is a two phase technique to efficiently compute exact, shortest paths.
It is not flexibile, i.e., $w_q=w_\ell$.
For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

@@

\subsection{Potentials using PHAST}

@@

The resulting algorithm is the basic CH-potential algorithm.
It is flexible and exact.
However, we can still improve the performance.
A quick investigation reveals, that a significant amount of running time is spent in the heuristic evaluation.
We therefore investigate methods to evaluate the heuristic less often.

@@

Heuristic evaluation is an expensive operation that we want to avoid.
The heuristic needs to be evaluated when a node is pushed into the queue.
By avoiding push operations, we can therefore avoid heuristic evaluations.
This is the objective of this section.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
In contrast, for road networks there also exist hierarchical speedup techniques, which do not need a heuristic.
They achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use such as Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
We call our technique CH-Potentials.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

@@

\centering
\includegraphics[width=\columnwidth]{fig/searchspace_st.png}

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
This is the tightest heuristic achievable during preprocessing.
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, it is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

In this section, we first describe Contraction Hierarchies, then PHAST, a CH extension, and finally CH-Potentials.

@@

A CH is a two phase technique to efficiently compute exact, shortest paths.
It is not flexible, i.e., $w_q=w_\ell$.
For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

@@

\subsection{PHAST based heuristic}

@@

The resulting algorithm is the basic CH-potential algorithm.
It is flexible and exact.
However, we can still improve the performance by reducing the number of queue operations and heuristic evaluations.

@@

Heuristic evaluation is an expensive operation that we want to avoid.
Together with the queue operations, it makes up a significant amount of the running time of A*.
The heuristic needs to be evaluated whenever a node is pushed into the queue.
As our experiments show, the total running time roughly corresponds to the number of queue pushs.
In this section, we propose optimizations to reduce the number of nodes pushed into the queue.
